biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-10-02 paper/PaperTRA/PaperTRA.tex
discussion + litrev

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can significantly replace private car trips.
This approach requires interventions and programs to make bicycling more
appealing, and the resulting public investments can have significant
social and environmental benefits.

According to the latest mobility survey conducted in 2018 {[}1{]}, the
LMA registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
number of intra-municipal trips --- with origin and destination in the
same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}3{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}4{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}5{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The mobility survey data {[}1{]} is the basis for this project and
defines the baseline scenario. Despite being conducted in the
pre-pandemic period (2017), this dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}6{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 {[}9{]} and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.} from
the World Health Organization. The HEAT tool provided estimates on the
shifting from car to cycling for a short term time horizon (i.e., one
year) and the long term (i.e., ten years). We considered two dimensions:
\emph{social} --- including the physical activity of cyclists, air
pollution exposure, and road casualties; \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}10{]} for a Tier 3 detail level. We used
a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
trips were considered to be made under urban conditions and at an
average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals. In
addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
{[}1{]}.

Regarding the PT, we considered the emission factor values reported in
the environmental and sustainability reports of the PT operators in the
LMA {[}11--14{]}.

@@

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}15{]} and the best up-to-date reference values for the
various gases {[}15--17{]}. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal {[}18{]}. We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%.

@@

Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
combined with PT trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment benefits are also presented, resulting from replacing
car trips with cycling.

@@

For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
car trips have the potential to be replaced by a combination of PT and
cycling (up to 5 km on bike).

\emph{Table \ref{tab:summary21} shows something that doesn't make
sense.}

@@

\emph{Say something about the avoided C02eq emissions in the first and
last legs (bike) is about 3/4 of the PT leg avoided emissions, which is
(expected?) and should not be overlooked when promoting the use of PT.
Making the accessibility to PT interfaces more atractive to cycling and
providing bicycle-friendly equipments (parking) can reduce more 75\%
CO2eq emissions than the shifted car to PT trip} (FM/RL help)

Our findings suggest that a bicycle and PT combination could viably
replace 20\% of current LMA trips, with an additional 12\% of PT
journeys prone to further substitution.

@@

The information on socio-economic benefits can support policy-makers in
prioritizing interventions to reduce the reliance on individual
motorized transportation, and better communicate their decisions by
providing the expected avoided GHG and air pollutant emissions and the
monetized socio-economic benefits for short and long terms. Different
routing profiles enable decision-makers to plan for different bicycle
user typologies and/or for different city cycling maturity levels
{[}19{]}.

@@

\hypertarget{acknowledgements}{%
\subsubsection*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{subsubsection}{Acknowledgements}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{INE:
\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
conselho de ministros n.º 131/2019,
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
to school in england, and associated health and carbon impacts:
Application of the {`}propensity to cycle tool{'}. Journal of Transport
\& Health. 12, 263--278 (2019).
https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Carris:
\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
de sustentabilidade 2019 - demostração não financeira}. {Carris -
Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Metropolitano de Lisboa:
\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{CP:
\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Transtejo:
\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
Sirtori, E., Vignetti, S., Del Bo, C.:
\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
to cost-benefit analysis of investment projects. Economic apraisal tool
for cohesion policy 2014-2020}. {European Commission - Directorate
General for Regional and Urban Policy} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
Hunt, A., Jong, G.D., Laird, J.:
\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
- Developing Harmonised European Approaches for Transport Costing and
Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
(2006).}

\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Nash, C., others:
\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
of accounts and marginal costs for transport efficiency: Final report}.
{Institute for Transport Studies, University of Leeds} (2003).}

\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
Rodoviária (ANSR)} (2021).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
Transportation Research Record. 2662, 125--133 (2017).
https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}

>>>>

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can significantly replace private car trips
{[}1, 2{]}. This approach requires interventions and programs to make
bicycling more appealing, and the resulting public investments can have
significant social and environmental benefits.

According to the latest mobility survey conducted in 2018 {[}3{]}, the
LMA registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
number of intra-municipal trips --- with origin and destination in the
same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}5{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}6{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}7{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The mobility survey data {[}3{]} is the basis for this project and
defines the baseline scenario. Despite being conducted in the
pre-pandemic period (2017), this dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}8{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively. Different routing profiles enable
decision-makers to plan for different bicycle user typologies and/or for
different city cycling maturity levels {[}11{]}.

@@

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 {[}12{]} and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.} from
the World Health Organization. The HEAT tool provided estimates on the
shifting from car to cycling for a short term time horizon (i.e., one
year) and the long term (i.e., ten years). We considered two dimensions:
\emph{social} --- including the physical activity of cyclists, air
pollution exposure, and road casualties; \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}13{]} for a Tier 3 detail level. We used
a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
trips were considered to be made under urban conditions and at an
average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals. In
addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
{[}3{]}.

Regarding the PT, we considered the emission factor values reported in
the environmental and sustainability reports of the PT operators in the
LMA {[}14--17{]}.

@@

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}18{]} and the best up-to-date reference values for the
various gases {[}18--20{]}. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal {[}21{]}. We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a discount rate of 5\% and inflation of
3\%.

@@

Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
combined with PT trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are also presented, resulting from
replacing car trips with cycling.

@@

For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
car trips have the potential to be replaced by a combination of PT and
cycling (up to 5 km on bike). \emph{Table \ref{tab:summary21} shows
something that doesn't make sense.}

@@

The emissions of CO\textsubscript{2}eq that are avoided during both the
initial and final journey segments account for about 75\% of the
emissions avoided during the PT segment. This finding, while expected --
due the zero cycling emissions, should not be overlooked when promoting
the PT use. Improving the safe accessibility to PT interfaces to
cyclists and providing bicycle-friendly amenities such as parking
facilities can potentially lead to a higher reduction in
CO\textsubscript{2}eq emissions, compared to a scenario where
individuals shift from car travel to car + PT combination.

The social impacts from replacing car trips to bicycle in first-and-last
legs, represent 95\% of the total socio-environmental benefits estimated
for all journey legs. Our findings suggest that a bicycle and PT
combination could viably replace 20\% of current LMA trips, with an
additional 12\% of PT journeys prone to further substitution.

@@

The information on socio-economic benefits can support policy-makers in
prioritizing interventions to reduce the reliance on individual
motorized transportation, and to better communicate their decisions by
providing the expected avoided GHG and air pollutant emissions and the
monetized socio-economic benefits for short and long terms.

@@

\hypertarget{acknowledgements.}{%
\subsubsection*{Acknowledgements.}\label{acknowledgements.}}
\addcontentsline{toc}{subsubsection}{Acknowledgements.}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Martens, K.: Promoting bike-and-ride: The dutch
experience. Transportation Research Part A: Policy and Practice. 41,
326--338 (2007).
https://doi.org/\href{https://doi.org/10.1016/j.tra.2006.09.010}{10.1016/j.tra.2006.09.010}.}

\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Rietveld, P.: The accessibility of railway stations: The
role of the bicycle in the netherlands. Transportation Research Part D:
Transport and Environment. 5, 71--75 (2000).
https://doi.org/\href{https://doi.org/10.1016/S1361-9209(99)00019-X}{10.1016/S1361-9209(99)00019-X}.}

\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{INE:
\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
conselho de ministros n.º 131/2019,
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
online system for sustainable transport planning. Journal of Transport
and Land Use. 10, (2017).
https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}

\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
to school in england, and associated health and carbon impacts:
Application of the {`}propensity to cycle tool{'}. Journal of Transport
\& Health. 12, 263--278 (2019).
https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}

\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
cyclists: Review of market segmentation methods for planning practice.
Transportation Research Record. 2662, 125--133 (2017).
https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Carris:
\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
de sustentabilidade 2019 - demostração não financeira}. {Carris -
Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Metropolitano de Lisboa:
\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{CP:
\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Transtejo:
\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
Sirtori, E., Vignetti, S., Del Bo, C.:
\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
to cost-benefit analysis of investment projects. Economic apraisal tool
for cohesion policy 2014-2020}. {European Commission - Directorate
General for Regional and Urban Policy} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
Hunt, A., Jong, G.D., Laird, J.:
\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
- Developing Harmonised European Approaches for Transport Costing and
Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
(2006).}

\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Nash, C., others:
\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
of accounts and marginal costs for transport efficiency: Final report}.
{Institute for Transport Studies, University of Leeds} (2003).}

\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
Rodoviária (ANSR)} (2021).}
